---
title: "Introduction to gemcox"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to gemcox}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 3.5)
```

GeM-Cox looks for latent subgroups of subjects whose **survival mechanisms**
differ: the same immune markers relate to the hazard differently in
different people. It combines a Gaussian mixture for clustering features
with a separate Cox model in each subgroup, all sharing one baseline hazard,
so subgroups differ only in their coefficient vectors. This vignette fits the
model to one simulated dataset, looks at what it estimates, and tests
whether two subgroups are needed.

```{r setup}
library(gemcox)
```

## Simulated data

`gemcox_simulate()` generates two equally likely latent clusters. Their
feature profiles are `mu_sep` apart and their Cox coefficient vectors are
`beta_sep` apart; the two coefficient vectors have equal length, so a
different mechanism is never confounded with a stronger effect. Here the
profiles overlap substantially (`mu_sep = 0.5`) and the mechanisms differ
clearly (`beta_sep = 2`).

```{r data}
d <- gemcox_simulate(n = 400, p = 6, mu_sep = 0.5, beta_sep = 2, seed = 42)
dim(d$X)
mean(d$status)          # proportion of subjects with an event
```

## Fitting the model

The defaults are the settings validated in the package simulations: K = 2,
`gamma = 1`, a small ridge penalty (`lambda = 0.05`, `alpha = 0`), a shared
baseline, diagonal covariance, and k-means initialisation.

```{r fit}
fit <- gemcox(d$X, time = d$time, status = d$status, K = 2)
fit
```

## Membership weights

Each subject gets a weight `tau` for each cluster rather than a hard label.
Here the weights are soft: most subjects are not confidently assigned.

```{r tau}
head(round(fit$tau, 3))
hist(apply(fit$tau, 1, max), breaks = 20, xlim = c(0.5, 1),
     main = "Largest membership weight per subject", xlab = "max tau")
```

This is expected. When subgroups differ mainly in mechanism, a single
subject's data carries little evidence about which subgroup they belong to,
and that evidence does not grow with the sample size. Individual labels
are therefore not the target of estimation; the **difference between the
subgroups' coefficient vectors** is.

## The two coefficient vectors

```{r coef}
round(coef(fit), 3)
contrast_hat  <- coef(fit)[, 1] - coef(fit)[, 2]
contrast_true <- d$beta_list[[1]] - d$beta_list[[2]]
## direction of the estimated contrast vs the truth (sign is arbitrary)
abs(sum(contrast_hat * contrast_true)) /
  sqrt(sum(contrast_hat^2) * sum(contrast_true^2))
```

`summary()` shows the coefficients next to their contrast, together with
mixing proportions, effective numbers of events per cluster and
convergence.

```{r summary}
summary(fit)
```

## Is a second subgroup needed?

`gemcox_heterogeneity_test()` tests K = 2 against K = 1, where the null
hypothesis is one mechanism with a shared coefficient vector. Its default
statistic is the likelihood ratio of the two fits, and its null
distribution comes from a parametric bootstrap under the fitted K = 1 model:
event times are resimulated from that model, censoring from its
Kaplan-Meier estimate, and both models are refitted each time.

The likelihood ratio requires `gamma = 1` (the default). For speed this
example uses only `B = 19` bootstrap replicates, which is enough to show
the call but gives a coarse p-value; use at least 200 in practice.

```{r test}
ht <- gemcox_heterogeneity_test(fit, B = 19, seed = 1)
ht
```

In the package simulations (n = 400-800, p = 10), this test:
- rejected at close to the nominal 5% under the null;
- had 30-57% power when the mechanisms differed by `beta_sep = 2`.

Those simulations used the earlier EM setting (`tol = 1e-4`,
`max_iter = 100`). Calibration at the current default (`tol = 1e-8`,
`max_iter = 1000`) has not yet been established, so treat these figures
as provisional. Power is modest with few events, so a non-significant
result is inconclusive: it does not show that the mechanisms are the
same.

## What gamma does

When GeM-Cox decides how strongly a subject belongs to each subgroup, it
combines two kinds of evidence: how the subject's markers compare with each
subgroup's typical profile, and how well each subgroup's survival model
explains what actually happened to the subject. `gamma` sets how much the
second kind of evidence counts.

* `gamma = 0` uses the markers only. Subgroups that differ in mechanism but
  not in marker profile are then invisible, however large the data.
* `gamma = 1` (the default) weights the outcome exactly as the joint
  probability model does. This is what lets the mechanism difference
  be recovered.
* Larger values make the outcome dominate and can create subgroups that
  are not there.

`gamma` is fixed at 1 rather than tuned: in simulation, data-driven choices
either pushed it to 0, losing the signal, or inflated false discoveries.

## Predicting for new subjects

A new subject's outcome is unknown, so their membership weights come from
their markers alone.

```{r predict}
new <- gemcox_simulate(n = 3, p = 6, mu_sep = 0.5, beta_sep = 2, seed = 7)
predict(fit, new$X, type = "tau")
predict(fit, new$X, type = "survival", times = c(50, 150, 300))
```

## When to use GeM-Cox

GeM-Cox is most useful when marker profiles overlap between subgroups but
are not identical, and survival mechanisms differ. Its advantage over a
single Cox model or a cluster-then-fit approach grows with the number of
events and shrinks as the number of features grows relative to the events.
Two subgroups (K = 2) is the validated setting.
